\begin{center}
\textbf{\large{Abstract}	}
\end{center}

\addcontentsline{toc}{chapter}{Abstract}
\begin{small}

Bacteriophages are viruses that infect bacteria, showing promise as alternatives to antibiotics in phage therapy. Accurate classification of phage lifestyle --- virulent or temperate --- is essential for therapeutic applications, as only virulent phages are safe for clinical use; temperate phages risk lysogenic conversion and horizontal transfer of virulence genes. Traditional laboratory-based methods are time-consuming and inapplicable to the majority of unculturable phages. Existing computational tools such as PhaTYP and DeePhage depend on gene prediction tools (Prodigal), which fail on short contigs from metagenomic data, while k-mer-based approaches suffer from information leakage.

This thesis proposes PhaBERT-CNN, a novel hybrid deep learning architecture combining the DNABERT-2 genome foundation model with an MKC (Multi-Kernel Convolutional) module. The MKC module consists of dual parallel branches: a CNN branch with kernel sizes 3, 5, and 7 for multi-scale local feature extraction, and a masked mean pooling branch for global sequence representation. PhaBERT-CNN applies Byte Pair Encoding (BPE) tokenization directly on raw DNA sequences without any gene prediction dependency. The model is trained with a two-stage strategy using discriminative learning rates. The dataset comprises 2,241 complete phage genomes (707 temperate, 1,534 virulent).

Experiments on four contig length groups --- Group A (100--400 bp), B (400--800 bp), C (800--1,200 bp), D (1,200--1,800 bp) --- with 5-fold stratified cross-validation demonstrate that PhaBERT-CNN achieves accuracies of 82.26\%, 87.38\%, 90.01\%, and 91.34\% respectively, improving 0.32--6.61\% over state-of-the-art methods including PhaTYP, DeePhage, and ProkBERT. The model maintains a strong balance between sensitivity (82.54\%--92.31\%) and specificity (81.30\%--87.71\%) across all length groups. The MKC module contributes a 0.74--4.09\% improvement over the DNABERT-2 baseline alone.

These results demonstrate the effectiveness of combining a genome foundation model with a task-specific architecture, particularly when operating directly on raw DNA sequences without gene prediction dependency. This represents a promising paradigm for biological sequence analysis tasks, especially metagenomic analysis involving incomplete or fragmented sequences.

\vspace*{1cm}
\textbf{Keywords}: \textit{Bacteriophage}, \textit{Lifestyle classification}, \textit{Deep learning}, \textit{BERT}, \textit{Convolutional neural networks}, \textit{Bioinformatics}, \textit{Metagenomics}
\end{small}
